---
title: "Will your code work in 2036?"
date: today
date-format: D MMMM YYYY
author: Richard J Telford
format: 
  revealjs:
    mermaid: 
      theme: forest
#mermaid-format: svg
editor: visual
code-annotations: hover
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

## 

Data + code !=
reproducible

. . .

Data + code + [environment]{style="color:red"} = reproducible

# Everything Changes,<br>your code changes not

## R changes

### R \< 3.6.0

```{r}
#| echo: true
#| eval: false
set.seed(1)
sample(LETTERS, size = 1)
```

```{r}
#| warning: false
RNGkind(sample.kind = "Rounding")
set.seed(1)
sample(LETTERS, size = 1)
```

::: fragment
### R \>= 3.6.0

```{r}
#| echo: true
#| eval: false
set.seed(1)
sample(LETTERS, size = 1)
```

```{r}
#| output-location: fragment
RNGkind(sample.kind = "default")
set.seed(1)
sample(LETTERS, size = 1)
```
:::

## Packages Change

```{r}
#| error: true
#| echo: true
#| output-location: fragment
library(ggplot2)
data(penguins, package = "palmerpenguins")

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_bar(position = "dodge") +
  opts(legend.position = "top")               # <1>
```

1.  `opts`? What are `opts`

::: fragment
### ggplot2 News June 2012

```         
ggplot2 0.9.2

----------------------------------------------------------------

THEME SYSTEM

* The `opts()` function has been renamed to `theme()`.
```
:::

## Solutions

```{r}
n_sol <- 4

d <- tibble(
  region = rep(c("Reproducibility", "Ease of use"), each = 3),
  x = c(0, 1, 1, 0, 0, 1) * n_sol, 
  y = c(0, 0, .9, 1, .1, 1)
    )

d |> ggplot(aes(x = x, y = y, fill = region)) +
  geom_polygon() +
  scale_x_continuous(expand  = c(0, 0)) +
  scale_y_continuous(expand  = c(0, 0)) +
  labs(y = NULL, x = "Solution", title = "Tradeoffs") +
  theme_minimal(base_size = 20) +
  theme(axis.text.y = element_blank(), 
        legend.position = "none", 
        panel.grid = element_blank()) +
  annotate(geom = "text", x = 0.9 * n_sol, y = 0.2, label = "Reproducibility", size = 10, hjust = 1) +
  annotate(geom = "text", x = 0.1 * n_sol, y = 0.8, label = "Ease of use", size = 10, hjust = 0) 

```

## Solution 0: name packages used

> values were analysed in R (R Core Team, 2017) using the qqman package (Turner, 2014).

## Citing R

```{r}
#| echo: true
# Citation for R
citation()
```

## Citing R packages

```{r}
#| echo: true
# Citation for an R package
citation("lme4")
```

## Solution 1: name & version of packages

> Numerical analyses were carried out using R v. 2.15.3 (R Core Team 2013) using the vegan package v. 2.0-9 (Oksanen et al. 2013).

Which packages to cite?

Potentially hundreds of packages used directly and indirectly.

## Finding package versions

``` markdown
Numerical analyses were carried out using R v. `{{r}} getRversion()`
(R Core Team 2013) using the vegan package v.`{{r}} packageVersion("vegan")`
(Oksanen et al. 2013).
```

Using `pakret` package

``` markdown
Numerical analyses were carried out using `{{r}} pkrt("R")`
using `{{r}} pkrt("vegan")`.
```

```{r}
options(pakret.render_mode = FALSE)
library(pakret)
```


Numerical analyses were carried out using `{r} pkrt("R")`
using `{r} pkrt("vegan")`.

## Solution 2: `renv` for reproducible R environment

-   detect which packages you use and their dependencies

-   keeps a **lockfile** - a list of the version of ALL the packages installed

-   makes a project library

-   will install correct version of packages in the lockfile when you start using a project

## Demo `renv`

1.  Make a new rstudio project.
2.  Run `renv::init()`
3.  Add an R script file with some `library()` calls
4.  Install packages with `renv::install()` (both CRAN and GitHub)
5.  Run `renv::status()`
6.  Run `renv::snapshot()` to save changes or `renv::restore()` to revert
7.  Repeat 3--6

## Using someone else's project with `renv`

1.  Clone the repo
2.  Open in RStudio
3.  use `renv::restore()` to install packages

## Pitfalls

-   `renv` only valid for the version of R it was built on
-   not all package can be discovered
    -   finds `library(dplyr)`, `tidyr::nest(...)` and dependencies

```{r}
#| eval: false
#| echo: true

tar_option_set(packages = c("tidyverse"))
```

-   solution use `targets::tar_renv()`

-   solution use R file with `library()` calls

-   Don't forget to `renv::update()` packages to get bugfixes.

## `renv` limitations

`renv` makes the list of packages available.

Does nothing for the all the external dependencies.

You want to reproduce a paper using R 3.6.3 ...

## Solution 3: Containers

Virtual machines vs containers

Containers

-   docker

-   podman

-   singularity/apptainer

Use for reproducibility, production, high-performance computers

## Install podman

Download and install podman from https://podman.io/

Test it is working:

```{.bash filename="Terminal"}
#| echo: true
podman run quay.io/podman/hello
```

## Docker images

Docker images are snapshots of a computer system

Can be downloaded and used

Images can be stacked

Images have version control

## Rocker - R images for Podman 

::::: columns
::: {.column width="20%"}
```{mermaid}
%%| label: rocker
%%| fig-width: 2

---
config:
  flowchart:
    padding: 10
---

flowchart BT
  ubuntu([ubuntu---])
  rver([rocker/r-ver-----])
  rstudio([rocker/rstudio------])
  tidyverse([rocker/tidyverse------])
  verse([rocker/verse-----])
  geospatial([rocker/geospatial------])
  geospatial-->verse-->tidyverse-->rstudio-->rver-->ubuntu

```
:::

::: {.column width="80%"}

```{.bash filename="Terminal"}
podman run --rm -ti -e PASSWORD=pass \
  -p 8787:8787 docker.io/rocker/rstudio:4.5.3
```
 `--rm` remove image after use\
 `-ti` interactive\
 `-e` environmental variables\
 `-p` port\
 `docker.io` registry\
 `rocker/rstudio:4.5.3` image & version tag

Open browser http://localhost:8787\
Usename: root\
Password: pass (or whatever you set)
:::
:::::

## Adding files

use `-v` option to mount files

```{.bash filename="Terminal"}
podman run --rm -ti -v .:/root -e PASSWORD=pass -p 8787:8787 \
  docker.io/rocker/rstudio
```

Files edited in container will be updated outside.

(Also possible to make files read-only)

## Adding dependencies with `renv` {.scrollable}

Make new image from a Dockerfile

```{.dockerfile filename="Dockerfile"}
# set base image
FROM docker.io/rocker/tidyverse:4.5.3 
# install renv into image
RUN R -e "install.packages('renv', repos = c(CRAN = 'https://cloud.r-project.org'))"
# copy renv lock into project & restore
WORKDIR /root
COPY renv.lock renv.lock
RUN R -e "renv::restore()"
```

Build Dockerfile into image with

```{.bash filename="Terminal"}
podman build -t thesis_image .
```

Run the image with 

```{.bash filename="Terminal"}
podman run -ti -v .:/root -p 8787:8787 thesis_image
```



## Podman compose {.scrollable}

Command line is too complex to want to write every time.

```{.yaml filename="compose.yml"}
services:
  rstudio:
    image: "docker.io/rocker/rstudio:4.5.3"
    ports:
      - "8787:8787"
    volumes:
      - type: bind
        source: .
        target: /root
    working_dir: /root
    environment:
      - PASSWORD=yourpassword

```

Run with

```{.bash filename="Terminal"}
podman compose compose.yml up

# close session with
podman compose down
```

## Podman desktop

Use for for starting, stopping and managing images.

## Solution 4: Nix - package manager

Nix language downloads packages

R interface with `rix` package

-   simplifies creating Nix expressions, which define reproducible environments.
-   lets you work interactively in RStudio


## Rix example

1) install Nix

2) Use rix to make Nix expression

```{r}
#| label: rix
#| echo: true
#| eval: false

library(rix)

rix(
  date = "2025-02-28", # set a date
  r_pkgs = c("dplyr", "ggplot2"), # List all the packages you need
  ide = "rstudio", # List whatever editor you need
  project_path = ".",
  overwrite = TRUE
)
```


3) Build nix environment

```{.bash filename="Terminal"}
nix-build
```



## Remaining problems

-   dependencies on external API (e.g. gbif, leaflet)
