---
title: "R and large data sets"
author: "Richard J. Telford"
institute: "University of Bergen"
editor: visual
format: revealjs
execute: 
  echo: true
---

## R and Memory

```{r}
#| label: options
#| echo: false
options(pillar.print_min = 5)
```

R loads objects into memory.

-   Data must be loaded into RAM, even if only part of it is needed.
-   Data must be stored in RAM, even if it is not used.
-   RAM is limited, data sets can be larger than the available RAM.

Large objects can cause the system to run out of memory.

Bad things happen if you run out of memory.

## Buy a better computer

-   RAM is expensive
-   Could never afford enough to deal with the largest datasets

## duckdb

Works with larger than memory data

Data processed on disk

`duckplyr` is a drop-in replacement for `dplyr`

Will read csv files and other formats

## Using `duckplyr`

```{r}
library(duckplyr)

penguins_file <- system.file("extdata/penguins.csv",
                             package = "palmerpenguins") 

# import from file
peng_duck <- read_csv_duckdb(path = penguins_file)

peng_duck
```

## Manipulate with `dplyr` verbs

```{r}
peng_duck |> 
  mutate(body_mass_g = as.numeric(body_mass_g)) |> 
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE), .by = species) |> 
  collect()

```

## `duckplyr`

Supports most `dplyr` function.

Will fall-back to `dplyr` if it has to.

-   `group_by()` fails - use `.by` argument instead

Supports many functions within `dplyr` verbs - each needs a translation

-   `sum()`, `mean()`, `n()`

## parquet files

Alternative to csv files

Compressed  - smaller file size

Import with `duckplyr` or `arrow` packages

## Databases

Useful for large, complex datasets.

```{r}
#| echo: false
tibble::tibble(site = rep(c(1, 2), 2), lat = rep(c(60.5, 62.3), 2), long = rep(c(5.1, 5.7), 2), elevation = rep(c(2300, 1500), 2), species = c("Poa alpina", "Poa alpina", "Geum rivale", "Geum rivale"), family = rep(c("Poaceae", "Rosaceae"), each = 2), functionGroup = rep(c("graminoid", "forb"), each = 2), abundance = c(5, 7, 14, 3)) |> 
  as.data.frame()
```

## Split into several tables

Sites

```{r}
#| echo: false

tibble::tibble(siteID = c(1, 2), lat = c(60.5, 62.3), long = c(5.1, 5.7), elevation = c(2300, 1500)) |> 
  as.data.frame()

```

Species

```{r}
#| echo: false

tibble::tibble(speciesID = c("PoaAlp", "GeuRiv"), species = c("Poa alpina", "Geum rivale"), family = c("Poaceae", "Rosaceae"), functionGroup = c("graminoid", "forb")) |> 
  as.data.frame()
```

Occurrences

```{r}
#| echo: false

tibble::tibble(siteID = rep(c(1, 2), 2), speciesID = c("PoaAlp", "PoaAlp", "GeuRiv", "GeuRiv"), abundance = c(5, 7, 14, 3)) |> 
  as.data.frame()
```

## Databases

Separate tables

-   Primary keys  - unique

-   Foreign keys

Stand alone database - SQLite, duckDB

Client-server databases - postgres

Connect with `DBI`

`dbplyr` makes a bridge to the database - no need to write your own SQL

## `DBI` + `dbplyr`

```{r}
#| eval: false
con <- DBI::dbConnect(RSQLite::SQLite(), here::here("database/seedclim.sqlite"))

sites_blocks_plots <- tbl(con, "sites") |> 
  inner_join(tbl(con, "blocks"), by = "siteID") |> 
  inner_join(tbl(con, "plots"), by = "blockID") |>  
  select(siteID, annual_precipitation_gridded, summer_temperature_gridded,
          blockID, plotID) |> 
  collect()
DBI::dbDisconnect(con)
```
